---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
library(here)
library(ggplot2)

#Reading the data
deaths_data <- read.csv(here("data/analysis_data/cleaned_file.csv"))
air_quality_data <- read.csv(here("data/analysis_data/cleaned_air_data.csv"))
avg_PM_conc_data <- read.csv(here("data/raw_data/avg PM2.5 conc data.csv"))
peak_PM_conc_data <- read.csv(here("data/raw_data/peak PM2.5 conc data.csv"))
avg_PM_conc_data <- avg_PM_conc_data %>% rename(Year = Category)
peak_PM_conc_data <- peak_PM_conc_data %>% rename(Year = Category)


# Assuming 'cleaned_air_data' is your air quality dataset and 'mortality_data' is your dataset of deaths.

library(tidyverse)
library(ggplot2)

# Ensure you have the correct column names by checking the names of your data frames
colnames(air_quality_data)
colnames(deaths_data)

# Merge the datasets
combined_data <- left_join(deaths_data, air_quality_data, by = "year")

head(combined_data)

# 2. Mortality Trend Plot for a single cause
# Repeat this plot changing the 'Cause' filter for each cause you are interested in.
ggplot(combined_data %>% filter(cause == 'All other forms of chronic ischemic heart disease'), 
       aes(x = year, y = total_deaths)) +  # Note the use of single quotes and backticks
  geom_line() +
  labs(title = "Trend in Deaths Due to Chronic Ischemic Heart Disease",
       x = "Year",
       y = "Total Deaths") +
  theme_minimal()

# 3. Relationship between Air Quality and Mortality Rates
# This could be a scatter plot with a line representing the trend.
ggplot(combined_data, aes(x = original_value, y = total_deaths, color = cause)) +  # Make sure the column name matches
  geom_point() +
  geom_smooth(method = "lm") +  # Add a linear trend line
  facet_wrap(~cause) +  # Create separate plots for each cause
  labs(title = "Relationship between Air Quality and Mortality Rates",
       x = "Air Quality Health Index",
       y = "Total Deaths") +
  theme_minimal()
```


# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.



Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)


Talk way more about it. 



# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.  

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.


### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.


# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```



# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}


# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows... 



## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```



\newpage


# References